\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-11-24 doc/publications/plos16/plos16.tex
Fix ref to chr1,2 and verbosify table names

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1~and~2), 
\marginpar{something wrong here. We do not report on chromosome 1 and 2.
Aidan} 
the time taken is excessive due to
the large amount of data being aggregated and processed by the single driver node during intermediate stages of building
the model (see Result~\ref{comp}). 
\marginpar{missing ref}
 This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In Table~\ref{table3}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
\end{tabular}
\begin{flushleft} Here the importance is the Gini impurity measure when building a model using the cell type as the label.
\end{flushleft}
\label{table3}
%\end{adjustwidth}
\end{table}

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
& Spark ML & NA & NA  \\
& ranger (C++)       &        NA     &        NA      \\
& ranger (C++)       &        NA     &        NA    \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
%\end{adjustwidth}
\end{minipage}
\end{table}

@@

Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{table10} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
here as a fraction of the number of variables) [Fig 1 a).

@@

\begin{table}[!ht]
\begin{adjustwidth}{-1.00in}{0in}
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees on 128 CPU cores with varying numbers of variables and samples.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
\cline{3-7}
\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
\hline                                                            
\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{table10}
\end{adjustwidth}
\end{table}

>>>>

Here we extend this toolkit to encompass supervised learning using random forests.  Spark ML's random forest
implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples
with only modest dimensionality.  Although Spark ML can build a random forest model on a subset of the data (chromosome
1), the time taken is excessive due to the large amount of data being aggregated and processed by the driver node
during intermediate stages of building the model (see Table~\ref{timingtable}). 
\marginpar{missing ref}
 This unbalanced work load where the driver node becomes the bottleneck and worker
nodes being idle prevents a seamless scaling to larger datasets. The memory requirements per executor also increases
with dimensionality due to the data types Spark ML uses, which we elaborate on in the discussion.

@@

In Table~\ref{featuretable}, we display the top 10 features, ranked by their Gini impurity measure in the random forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Features from the TCGA BRCA dataset ranked by their importance.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
{\bf Variant} & {\bf Gene} & {\bf Importance}\\ \thickhline
x:xxxx & cell2 row 1 & cell3 row 1\\ \hline
x:xxxx & cell2 row 2 & cell3 row 2\\ \hline
x:xxxx & cell2 row 3 & cell3 row 3\\ \hline
\end{tabular}
\begin{flushleft} Here the importance is the Gini impurity measure when building a model using the cell type as the label.
\end{flushleft}
\label{featuretable}
%\end{adjustwidth}
\end{table}

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX  & 7hr 14min \\
& Spark ML & NA & NA  \\
& ranger (C++)       &        NA     &        NA      \\
& ranger (C++)       &        NA     &        NA    \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{timingtable}
%\end{adjustwidth}
\end{minipage}
\end{table}

@@

Firstly we look at CursedForest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying \mtry\ fraction and the number of CPU cores allocated to the execution. The raw results are
presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synth} below. Regardless of the number of cores used,
\cursedforest\ displays approximately linear dependency between the execution time and the \mtry\ parameter (presented
here as a fraction of the number of variables) [Fig 1 a).

@@

\begin{table}[!ht]
\begin{adjustwidth}{-1.00in}{0in}
\centering
\caption{
{\bf Time in seconds to build a random forest of 100
  trees on 128 CPU cores with varying numbers of variables and samples.}}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{2}{|l|}{\multirow{2}{*}{}}           & \multicolumn{5}{c|}{Number of variables} \\
\cline{3-7}
\multicolumn{2}{|l|}{}                               & \bf{150,000} & \bf{500,000} & \bf{2,500,000}  & \bf{10,000,000} & \bf{50,000,000} \\
\hline                                                            
\multirow{4}{*}{Number of samples}      & \bf{1000}  & 21.6  & 26.7  & 59.8  & 133.5  & 668.8 \\
                                        & \bf{5000}  & 45.7  & 71.9  & 186.2 & 729.3  & 2887.4 \\
                                        & \bf{10000} & 70.2  & 120.2 & 397.3 & 2237.8 & 5747.8 \\
\hline
\end{tabular}
\begin{flushleft} 
\end{flushleft}
\label{synthetictimingtable}
\end{adjustwidth}
\end{table}
